Modelling last glacial cycle ice dynamics in the Alps

### 2017-07-04 articles/cycle/alps-cycle.tex
Update notes and todos to reflect new plans.

    \note{For the Nature Geoscience format this abstract will need to be
          converted to an ``introductory paragraph''.}

@@

    \note{An important assumption in the method is that the MIS~4 glaciation
          was less (or equally) extensive than the MIS~2, i.e. that the LGM
          outline from \citet{Ehlers.etal.2011} dates indeed from the LGM
          (MIS~2). A different assumption could be that this outline represents
          the most extensive glaciation in the last 120\,ka. I we allow the
          MIS~2 glaciation to be a bit smaller than MIS~4, simulations driven
          by GRIP and MD01-2444 also give reasonable results.}

@@

    \idea{Matthias: The selection of figures in the main text and SOM is good.
          I find Figure 2 quite difficult to understand at first glance. I
          would combine it (already before submission) with Fig. 3. the Figure
          can then focus both on the temporal advance/retreat patterns of the 4
          lobes (four panels on the left side of figure) and the spatial
          differences in max. ice elevation (1 panel on right side of figure,
          with centerlines indicated). I would have the one with trimline
          elevation again in the Supplementary.}

@@

    \begin{figure}
      \centerline{\includegraphics{alpcyc_hr_inputs}}
      \caption{%
        \textbf{(a)} July mean near-surface air temperature and
        \textbf{(b)} January precipitation from WorldClim
        \citep[1960--1990]{Hijmans.etal.2005}, and
        \textbf{(c)} Mordern July standard deviation of daily mean temperature
        from the ERA-Interim \citep[1979--2012]{Dee.etal.2011} from the
        reference monthly climatology used to force the surface mass balance
        (PDD) component of the ice sheet model.
        \textbf{(d)} Geothermal heat flow from applying the similarity method
        to multiple geophysical proxies \citep{Goutorbe.etal.2011} used as a
        boundary condition to the bedrock thermal model 3\,km below the
        ice-bedrock interface.
        \idea{Since this will be in the supplement, I can also add additional
              information such as winter temperature, summer precipitation,
              winter PDD SD, and/or basal topography with modern glaciers
              removed, e.g. in the Bernese Alps.}}
      \label{fig:inputs}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_lr_timeseries}}
      \caption{%
        \textbf{(a)} Temperature offset time-series from ice core and ocean
        records (Table~\ref{tab:records}) used as palaeo-climate forcing for
        the ice sheet model.
        \textbf{(a)} Modelled total ice volume through the last 120\,ka,
        expressed in meters of sea level equivalent (m~s.l.e.). Gray fields
        indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2 and
        MIS~4 according to a~global compilation of benthic \chem{\delta^{18}O}
        records \citep{Lisiecki.Raymo.2005}.}
      \label{fig:timeseries}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_lr_footprints}}
      \caption{%
        \textbf{(a--c)} Cumulative extent of modelled ice cover during MIS~2
        (29--14\,ka) using temperature time-series scaling factors
        (Table~\ref{tab:records}) adjusted to obtain model results in agreement
        with the Last Glacial Maximum (LGM) geomorphological reconstruction
        \citep[solid red line,][]{Ehlers.etal.2011}.
        \textbf{(d--f)} Cumulative extent of modelled ice cover during MIS~4
        (71--57\,ka). Only the simulation driven by the EPICA temperature
        time-series yields reasonable MIS~4 ice cover.}
      \label{fig:footprints}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
      \caption{%
        \textbf{(a)} Modelled bedrock topography (grey) ice surface topography
        (200\,m contours) and ice surface velocity (blue) in the Alps
        21~thousand years (ka) before present. Modelled Last Glacial Maximum
        (LGM) ice extent (dashed orange line) and geomorphological
        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}. The
        background map consists of depressed SRTM \citep{Jarvis.etal.2008}
        topography and Natural Earth Data \citep{Patterson.Kelso.2015}.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model \citep[black
        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
        last glacial cycle (120--0\,ka), expressed in meters of sea level
        equivalent (m~s.l.e., blue curve). Gray fields indicate Marine
        Oxygen Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
        a~global compilation of benthic \chem{\delta^{18}O} records
        \citep{Lisiecki.Raymo.2005}.}
      \label{fig:lgmvel}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_hr_timing}}
      \caption{%
        \textbf{(a)} Timing of the Last Glacial Maximum (LGM) given by the
        modelled age (colour mapping) and value (200 m contours) of maximum
        surface elevation throughout the entire simulation.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model (black curve),
        and modelled glaciated area around the LGM (coloured curve). The LGM
        is here modelled as a time-transgressive event.
        \note{The Nature Geoscience format is limited to four figures. Now we
              indeed have four figures, albeit large ones.
              Figure~\ref{fig:timing} will interest the dating people but is
              not crucial for the paper. My plan is to include in the main text
              for the first submission and be prepared that the editor may want
              to move it into supplementary material.}}
      \label{fig:timing}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_hr_profiles}}
      \caption{%
        Modelled extent of glaciation along selected profiles for the Lyon,
        Solothurn, Rhine and Ivrea glaciers.
        \todo{Implement extraction of ice thickness along a more detailed
              profile, and add the other glaciers.}
        \idea{We can also move this as panel b on Fig.~\ref{fig:timing}, with
              profiles drawn on panel a. But then we should probably limit it
              to two glaciers instead of four.}}
      \label{fig:profiles}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_hr_erosion}}
      \caption{%
        \textbf{(a)} Modelled total erosion integrated in time over the entire
        simulation (120--0\,ka) is highest in the deep valleys and cirques of
        the western Alps.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model (black curve),
        and modelled erosion rate integrated in space over the entire
        ice-covered area. The local erosion rate, $\dot{e}$, is computed from
        the sliding velocity, $\vec{v}_{\mathrm{b}}$, through
        $\dot{e} = K_g \cdot |\vec{v}_{\mathrm{b}}|^{l}$, with
        $l = 2.02$ and $K_g = 2.7\e{-7}\,m^{1-l}\,a^{l-1}$
        \citep{Herman.etal.2015}.}
      \label{fig:erosion}
    \end{figure}

@@

    \begin{table*}
      \caption{%
        Parameter values used in the ice sheet model.
        \idea{If I need to re-run all simulations, we could rethink bedrock,
              lithosphere and PDD parameters.}}
      \label{tab:params}
      \noindent\small\makebox[\textwidth]
      {\begin{tabular}{llrll}
        \toprule

@@

    \begin{table*}
      \caption{%
        Palaeo-temperature proxy records and scaling factors yielding
        temperature offset time-series used to force the ice sheet model
        through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
        corresponds to the scaling factor adopted to yield Last Glacial Maximum
        ice limits in the vicinity of mapped end moraines
        (Fig.~\ref{fig:footprints}a), and $[{\Delta}T_{\textrm{TS}}]_{32}^{22}$
        refers to the resulting mean temperature anomaly during the period 32
        to~22\,\unit{ka} after scaling.
        \note{The EPICA scaling factor is close to one, and that ended up
              yielding too much ice in the high-resolution simulation
              (Fig.~\ref{fig:lgmvel}), though not so at low resotion
              (Fig.~\ref{fig:footprints}b). If I re-run the whole thing, I will
              probably fix it to one (i.e. use EPICA temperature anomalies
              without applying any scaling).}}
      \label{tab:records}
      \noindent\small\makebox[\textwidth]
      {\begin{tabular}{lccccccl}
        \toprule

>>>>

@@

    \note{An important assumption in the method is that the MIS~4 glaciation
          was less (or equally) extensive than the MIS~2, i.e. that the LGM
          outline from \citet{Ehlers.etal.2011} dates indeed from the LGM
          (MIS~2). A different assumption could be that this outline represents
          the most extensive glaciation in the last 120\,ka. If we allow the
          MIS~2 glaciation to be a bit smaller than MIS~4, simulations driven
          by GRIP and MD01-2444 also give reasonable results.}

@@

    \idea{Matthias: The selection of figures in the main text and SOM is good.
          I find Figure 5 quite difficult to understand at first glance. I
          would combine it (already before submission) with Fig. 6. the Figure
          can then focus both on the temporal advance/retreat patterns of the 4
          lobes (four panels on the left side of figure) and the spatial
          differences in max. ice elevation (1 panel on right side of figure,
          with centerlines indicated). I would have the one with trimline
          elevation again in the Supplementary.}

    \idea{Add trimline figure in the main text now.}

@@

    \begin{figure}
      \centerline{\includegraphics{alpcyc_hr_inputs}}
      \caption{%
        \textbf{(a)} July mean near-surface air temperature and
        \textbf{(b)} January precipitation from WorldClim
        \citep[1960--1990]{Hijmans.etal.2005}, and
        \textbf{(c)} Mordern July standard deviation of daily mean temperature
        from the ERA-Interim \citep[1979--2012]{Dee.etal.2011} from the
        reference monthly climatology used to force the surface mass balance
        (PDD) component of the ice sheet model.
        \textbf{(d)} Geothermal heat flow from applying the similarity method
        to multiple geophysical proxies \citep{Goutorbe.etal.2011} used as a
        boundary condition to the bedrock thermal model 3\,km below the
        ice-bedrock interface.
        \todo{Fix geothermal plot.}}
      \label{fig:inputs}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_lr_timeseries}}
      \caption{%
        \textbf{(a)} Temperature offset time-series from ice core and ocean
        records (Table~\ref{tab:records}) used as palaeo-climate forcing for
        the ice sheet model.
        \textbf{(a)} Modelled total ice volume through the last 120\,ka,
        expressed in meters of sea level equivalent (m~s.l.e.). Gray fields
        indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2 and
        MIS~4 according to a~global compilation of benthic \chem{\delta^{18}O}
        records \citep{Lisiecki.Raymo.2005}.
        \todo{Add temporary subfigure labels. Refine temperature offsets.}}
      \label{fig:timeseries}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_lr_footprints}}
      \caption{%
        \textbf{(a--c)} Cumulative extent of modelled ice cover during MIS~2
        (29--14\,ka) using temperature time-series scaling factors
        (Table~\ref{tab:records}) adjusted to obtain model results in agreement
        with the Last Glacial Maximum (LGM) geomorphological reconstruction
        \citep[solid red line,][]{Ehlers.etal.2011}.
        \textbf{(d--f)} Cumulative extent of modelled ice cover during MIS~4
        (71--57\,ka). Only the simulation driven by the EPICA temperature
        time-series yields reasonable MIS~4 ice cover.
        \todo{Refine temperature offsets. Try to make reduced precipitation
              outlines more visible.}}
      \label{fig:footprints}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
      \caption{%
        \textbf{(a)} Modelled bedrock topography (grey) ice surface topography
        (200\,m contours) and ice surface velocity (blue) in the Alps
        21~thousand years (ka) before present. Modelled Last Glacial Maximum
        (LGM) ice extent (dashed orange line) and geomorphological
        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}. The
        background map consists of depressed SRTM \citep{Jarvis.etal.2008}
        topography and Natural Earth Data \citep{Patterson.Kelso.2015}.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model \citep[black
        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
        last glacial cycle (120--0\,ka), expressed in meters of sea level
        equivalent (m~s.l.e., blue curve). Gray fields indicate Marine
        Oxygen Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
        a~global compilation of benthic \chem{\delta^{18}O} records
        \citep{Lisiecki.Raymo.2005}.
        \note{I expect the simulation to complete in late July.}}
      \label{fig:lgmvel}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_hr_timing}}
      \caption{%
        \textbf{(a)} Timing of the Last Glacial Maximum (LGM) given by the
        modelled age (colour mapping) and value (200 m contours) of maximum
        surface elevation throughout the entire simulation.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model (black curve),
        and modelled glaciated area around the LGM (coloured curve). The LGM
        is here modelled as a time-transgressive event.
        \note{I expect the simulation to complete in late July.}}
      \label{fig:timing}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_hr_profiles}}
      \caption{%
        Modelled extent of glaciation along selected profiles for the Lyon,
        Solothurn, Rhine and Ivrea glaciers.
        \todo{Implement extraction of ice thickness along a more detailed
              profile, and add the other glaciers.}
        \idea{We can also move this as panel b on Fig.~\ref{fig:timing}, with
              profiles drawn on panel a. But then we should probably limit it
              to two glaciers instead of four.}
        \note{I expect the simulation to complete in late July.}}
      \label{fig:profiles}
    \end{figure}

    \begin{figure}
      \centerline{\includegraphics{alpcyc_hr_erosion}}
      \caption{%
        \textbf{(a)} Modelled total erosion integrated in time over the entire
        simulation (120--0\,ka) is highest in the deep valleys and cirques of
        the western Alps.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model (black curve),
        and modelled erosion rate integrated in space over the entire
        ice-covered area. The local erosion rate, $\dot{e}$, is computed from
        the sliding velocity, $\vec{v}_{\mathrm{b}}$, through
        $\dot{e} = K_g \cdot |\vec{v}_{\mathrm{b}}|^{l}$, with
        $l = 2.02$ and $K_g = 2.7\e{-7}\,m^{1-l}\,a^{l-1}$
        \citep{Herman.etal.2015}.
        \note{I expect the simulation to complete in late July.}}
      \label{fig:erosion}
    \end{figure}

@@

    \begin{table*}
      \caption{%
        Parameter values used in the ice sheet model.
        \todo{Update parameters to new configuration.}}
      \label{tab:params}
      \noindent\small\makebox[\textwidth]
      {\begin{tabular}{llrll}
        \toprule

@@

    \begin{table*}
      \caption{%
        Palaeo-temperature proxy records and scaling factors yielding
        temperature offset time-series used to force the ice sheet model
        through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
        corresponds to the scaling factor adopted to yield Last Glacial Maximum
        ice limits in the vicinity of mapped end moraines
        (Fig.~\ref{fig:footprints}a), and $[{\Delta}T_{\textrm{TS}}]_{32}^{22}$
        refers to the resulting mean temperature anomaly during the period 32
        to~22\,\unit{ka} after scaling.
        \todo{Correct and refine temperature offsets. Add reduced precipitation
              runs.}}
      \label{tab:records}
      \noindent\small\makebox[\textwidth]
      {\begin{tabular}{lccccccl}
        \toprule
